# Exam  {#L22-exam}

## About the exam

First, the details of the exam:

* Check your exam timetable for the date and time of the exam. For most people (without unusual clashes or special arrangements), the exam will happen on **Monday 16 January at 1400**.
* The exam will happen in person on campus. For most people this will be in the Sports Hall.
* The exam will last for 2 hours (except for some students with special arrangements).
* You are allowed to use a basic non-programmable calculator for the exam.
* The exam will be "closed-book" style: you are not permitted to bring notes into the exam hall.
* The exam will contain multiple-choice questions, short questions, and long questions. See below for more details about the structure of the exam.
* A page of [statistical tables for the normal distribution](https://mpaldridge.github.io/math1710/stat-tab.pdf) will be attached to the exam paper. There will not be a formula book.
* The exam makes up 70% of your mark for this module. The pass mark for this module is 40%. You must pass the exam to pass the module. The pass mark for the exam is also 40%, or 32 marks out of 80.

The exam will be in three sections:

* **Section A** contains 10 multiple-choice questions, worth 2 marks each, for a total of 20 marks. These each require a single letter answer. No rough work will be read or marked. You will enter these letter answers on a multiple-choice "bubble sheet"; there is an example of this sheet on Minerva. (Recommended time: 3 minutes per question; 30 minutes total.)
* **Section B** contains 10 single-part "short answer" questions, worth 2 marks each, for a total of 20 marks. These require a clear answer and brief working or explanation. Typically, a correct answer with clear working will get 2 marks; partial progress or a correct answer with totally inadequate working will get 1 mark. You will answer these in an answer booklet. (Recommended time: 3 minutes per question; 30 minutes total.)
* **Section C** contains 2 multi-part "long answer" questions in multiple parts, worth 20 marks each, for a total of 40 marks. Parts of these questions require full detailed answers, as in assessed work from problem sheets, and will have marks set aside for quality of explanations. You will answer these in the same answer booklet as Section B. (Recommended time: 30 minutes per question; 1 hour total.)

## Past papers

School of Mathematics policy is that for Level 1 modules, the three most recent past papers (that used the same exam format as your exam) are made available. For MATH1710, these are

* The 2021--22 paper (from January 2022), available only on Minerva (in the "Exam" folder for this module).
* Not the 2020--21 paper, as this was held in pandemic times, and had a different format.
* The 2019--20 (January 2020) and 2018--19 (January 2019) papers, which are available on [the University past papers website](https://students.leeds.ac.uk/exampapers?action=search&keyword=MATH1710).
* (At the time of writing, it appears that the 2017--18 (January 2018) paper is also on [the University past papers website](https://students.leeds.ac.uk/exampapers?action=search&keyword=MATH1710), so it seems you have a extra bonus past paper!)

(Your username on [the University past papers website](https://students.leeds.ac.uk/exampapers?action=search&keyword=MATH1710) is just the `xy22abc` part of your username, without the `@leeds.ac.uk`.)

University policy is that we are not permitted to give out complete worked solutions to past exam papers. Instead, I have provided a "checksheet" for the three past papers 2021-22, 2019--20 and 2018--19. This should give enough information for you to check your own answers, if you use the papers for your revision, but is rather bare-bones and doesn't give a lot of detail.

I strongly recommend using all these past papers as part of your revision, and resisting the temptation to peek at the checksheet before trying the questions yourself, perhaps multiple times. You may wish to test yourself by trying one of the papers sight-unseen under exam-like conditions.

Some notes on the past papers:

* 2021-22 paper:
  * This is the only past paper written by me. Therefore, it's the only past paper based on the module (almost) exactly as taught this year, and using the styles of questions I tend to write. For this reason, I consider this paper a valuable resource for your revision, and one you should not waste by too-quickly looking at the checksheet.
  * That said, I think I erred by making this exam too hard. I have written your exam to be easier -- I hope an average student should get about 8 more marks (10% more) on your exam than on this past paper. You should bare this in mind when assessing your performance on this past paper.
* The two older past papers (2019--20 and 2018--19):
  * These papers write $\Pr(A)$ for probabilities, where we write $\mathbb P(A)$, and write $\mathrm{E}[X]$ for expectations, where we write $\mathbb EX$.
  * These papers sometimes write $\subseteq$ for "is a subset of", where we write $\subset$.
  * In R questions, often `=` is used for assignment, as in `variable = 2 + 3`, whereas we prefer to use `<-`, as in `variable <- 2 + 3`.
  * Although these papers in Section A instructed students to "write down a single letter", you will fill in your multiple-choice answers on the "bubble sheet" provided.
* 2018--19 paper:
  * Questions A6 and B9: This writes $\Omega_X$ for the range (or "range space") of a random variable $X$, where we wrote $\operatorname{Range}(X)$.
  * Question C2(b): Some of the notation here is a bit different to ours; for example, writing $l(x|p)$ for the likelihood.
* 2019--20 paper:
  * Question B9: You are not expected to be able to answer this question.
  * Question C2(b): Some of the notation here is a bit different to ours; for example, writing $l(x|p)$ for the likelihood.
  
<!---
* 2020--21 paper:
  * Because this was a "take-home" paper, there was no multiple-choice section, which it was thought would be too easy for nefarious students to cheat on. Instead, there were twice as many short questions.
  * Because this was an "open-book" paper, there were fewer "fact" questions than usual, such as stating definitions or giving standard proofs from the notes.
  * Question A11(ii): You are not expected to be able to answer this question.
  * Question A12: An event $A$ is said to be "preferable" for an event $B$ is $\mathbb P(B \mid A) > \mathbb P(A)$. You are not expected to know this definition (although, once told the definition, you should be able to work with it).
  * Question A18: You are not expected to be able to answer this question.
--->


## Exam FAQs

### Calculators

::: {.thpart}
*Can I use a calculator in the exam?*
:::

::: {.thpart}
*What sort of calculator can I use in the exam?*
:::

You may use a calculator for MATH1710 exam. The School of Mathematics rule is this:

> You may use a *simple, non-programmable* calculator.

You may feel that this rule is unhelpfully vague. I am not totally without sympathy for this view, but this is nonetheless the School rule. The Maths Student Support Office did recently offer the following extra guidance:

> The only rule is that, where calculators are allowed in an exam, any basic (i.e. non-programmable) calculator is allowed. Calculators that can solve calculus problems or do matrix calculations are allowed, as long as they are non-programmable. Calculators are only allowed in certain exams, and the module leader will be able to tell you whether a calculator is allowed on their exam.
>
> If you are not sure whether your calculator is non-programmable then check the manufacturer's website. Any calculator that is marketed as "programmable" or "graphic" is unlikely to be allowed in an exam. Examples of calculators that are allowed include the Casio FX-85GTX and the Casio FX991-EX. 
>
> Please note: the school does not provide a definitive list of approved and non-approved calculators. We do not verify / pre-approve calculators (some schools may provide verification stickers but these are not needed for MATH exams). It will be the exam invigilators who will check calculators during the examination."

I notice that the Casio FX-85GTX is currently available for [£14 from WHSmith](https://www.whsmith.co.uk/products/casio-fx85gtx-black-scientific-calculator/4549526607844.html) or [£15 from Amazon](https://www.amazon.co.uk/dp/B07L68RQCX).

### Revision

::: {.thpart}
*What tips do you have for revising for a University maths exam?*
:::

I always feel a bit self-conscious answering this question -- I last did a maths exam 15 years ago, while many of you last did a maths exam six or seven months ago. Surely it's *you* who should be giving *me* revision advice!

That said, I like [these exam revision tips from my former colleague Dr Oliver Johnson](https://people.maths.bris.ac.uk/~maotj/notes/revise.png), who teaches the equivalent module to MATH1710 at the University of Bristol. I particularly endorse Dr Johnson's points 1 and 2:

> 1. **Triage** your notes into: stuff you know already, stuff you almost understand, stuff you'll never understand. Focus your work on the middle third (this may change over time).

> 2. **Revision is an active process:** summarise your notes, solve problems, test yourself. If you just read, you'll be staring into space 10 minutes later.

Read the rest of Dr Johnson's revision tips [here](https://people.maths.bris.ac.uk/~maotj/notes/revise.png).

### R questions on the exam

::: {.thpart}
*Can there be R questions on the exam?*
:::

Yes!

To state the obvious, you will be sitting in an exam hall with a paper booklet and a pen, so I can't ask you to read a 400-row CSV file into RStudio and tell me the standard deviation of the fourth column. But there are two types of questions you could be asked.

The first type is "What would the output from R be if you entered the code...". An example of this from the 2021-22 paper is this multiple-choice Section A question:

:::: {.thpart}
**A1.** What would the output from the following R code be?

```{r, eval = FALSE}
data <- 1:10
round(sd(data), digits = 2)
```

**A** 3.03, **B** 5.5, **C** 6.36, **D** 9.17, **E** 40.5
::::

To answer this, you will need to know that `1:10` means the dataset $(1, 2, \dots, 10)$, and that the second line calculates the standard deviation of that dataset to 2 decimal places. You could then calculate that standard deviation to get the answer 3.03, which is **A**. Alternatively, you could notice that the answer must be at most 4.5 (the maximum distance from any data point $x_i$ to the mean $\bar x = 5.5$), and so **A** 3.03 is the only plausible answer.

The second type is: "Write down some R code that would calculate...". Here's an example of a Section B question from a past paper in this style:

::: {.thpart}
**B4.** In RStudio, a vector `x` contains the values in the range of a discrete random variable $X$ and a vector `p` contains the corresponding probabilities. Write down an R command to evaluate $\mathbb EX(X-1)$ from `x` and `p`.
:::

By the law of the unconscious statistician, this is calculated by
\[ \mathbb EX(X-1) = \sum_x x(x-1)\,p(x) . \]
This can be calculated with the R code

```{r, eval = FALSE}
sum(x * (x - 1) * p)
```

(See also the discussion of [discrete random variables in R from the last lecture](#faq-R-RV).)

I can reveal that your exam will have two questions (or part-questions) on R, totalling 5 of the 80 marks.


### Statistical tables

### How much should I write?




<!--
## Revision

It seems a bit strange for me to be giving you revision advice -- I last did a maths exam more than a decade ago, while you last did a maths exam a few months ago!
-->