# Problem Sheet 4  {#P4 .unnumbered}




<style>
.myanswers { 
display: none !important;
}
</style>



\commfalse


::: {.mysummary}
This is Problem Sheet 4. This problem sheet covers [Section 6](#S06-discrete-dist), and next week will have questions added on Section 7. You should work through all the questions on this problem sheet so far during Week 6, in preparation for your tutorial in Week 8. The problem sheet contain an assessed question, to which a second will be added next week, and which are due in by **2pm on  Monday 22 November**.

Please also fill in the [mid-semester survey](https://forms.office.com/Pages/ResponsePage.aspx?id=qO3qvR3IzkWGPlIypTW3ywARQdZlKXRHsLcXi_ngX8NUOVVWTlQ3SVZHNjJWQVcwTFU0RDlaOFk0Ni4u) if you haven't done so yet.
:::

## A: Short questions {#P4-short .unnumbered}

```{r, child = "problems/questions/P4-short.Rmd"}
```

## B: Long questions {#P4-long .unnumbered}

```{r, child = "problems/questions/P4-long.Rmd"}
```


## C: Assessed questions {#P4-assessed .unnumbered}

::: {.mysummary}
The last two questions are **assessed questions**. These two questions count for 3% of your final mark for this module.

The deadline for submitting your solutions is **2pm on Monday 22 November** at the beginning of Week 9, although I strongly recommend completing and submitting your work during Week 8. Submission will be via Gradescope; submission will open on Monday 15 November.
Your work will be marked by your tutor and returned on Monday 29 November, when solutions will also be made available.

Both questions are "long questions", where the marks are not only for mathematical accuracy but also for the clarity and completeness of your explanations. 

You should not collaborate with others on the assessed questions: your answers must represent solely your own work. The University's rules on [academic integrity](https://library.leeds.ac.uk/info/1401/academic_skills/46/academic_integrity_and_plagiarism) -- and the related punishments for violating them -- apply to your work on the assessed questions.
:::

```{r, child = "problems/questions/P4-assessed.Rmd"}
```

## Solutions to short questions {#P4-short-sols .unnumbered}

**A1.** 


