\commtrue

# Solutions and group feedback {#solutions .unnumbered}

::: {.mysummary}
There are many ways you get feedback on this module, both group feedback (feedback that is generally relevant to many people in the class) and individual feedback (feedback based specifically on your own approach to the work).

* You will have got both individual and group feedback in your tutorial (the more you speak up in your tutorial, the individual the feedback you get in return).
* These solutions include more group feedback on common issues for the class.
* Most importantly, when your work on assessed questions is marked, individual feedback will be given via the Gradescope site, and it is very important that you read that feedback.
* Finally, students who would like even more feedback can discuss their work in the "office hours" drop-in sessions.
:::

## Problem Sheet 1 {#P1-solutions .unnumbered}

```{r, child = "problems/questions/P1-short.Rmd"}
```

```{r, child = "problems/questions/P1-long.Rmd"}
```