---
title:  "R Worksheet 1: R Basics"
author: "MATH1710 Probability and Statistcs I"
date:   "University of Leeds, 2021--22"
output: html_document
---

*This R worksheet does not include any assessed questions.*

------------------------------------------------------------------------

# How to use R and RStudio

This worksheet assumes you or using a computer with the programming language R and the program RStudio installed. This might be:

* a University computer, which all have R and RStudio installed;
* your own computer on which you have first installed R and then installed RStudio, following [the instructions in the notes](https://mpaldridge.github.io/math1710/R.html#r-install);
* using the [RStudio Cloud](https://rstudio.cloud/).

Open RStudio. To enter commands, you should type them into the "Console" -- this normally takes up either the left half of the screen, or the bottom-left quarter. When you open RStudio, there will be some information about R in the console, in black writing, then an arrow `>` in blue. You can type commands next to this blue arrow.

> **Exercise 1.1.** Type `2 + 3` into the console next to the arrow, and then press Enter. What happens?

In these worksheets, R commands for you to enter will look like this

```{r, eval = FALSE}
2 + 3
```

We don't write down the `>` in the worksheets, because you don't have to type the `>` either. But remember that you always type your commands next to the blue arrow.

In these worksheets, the information returned by R will look like this:

```{r, echo = FALSE}
2 + 3
```

The double hash-sign `##` will not appear on your screen. (We'll explain later why we use them here.) The `[1] 5` is what you should have seen as R's response in Exercise 1.1. First, the `[1]` just tells you that what follows is the first part of the answer -- since our answer here only has one part, we can ignore this. (It is useful when the answer is a very long vector or a table of values.) Then the `5` is the answer we want.

## Using R as a calculator

The simplest way to use R is to use it to perform simple calculations for us -- like the `2 + 3` we saw earlier. Similarly, R can perform subtraction `-`, multiplication `*`, division `\` and powers `^`. So, for example, we can find $(4 + 5) \times 6$ by entering into the Console:

```{r}
(4 + 5) * 6
```

Spaces (with some exceptions) are ignored by R, so `2+3`, `2 + 3` and `2    +3` all work equally well -- you can use whichever you find easiest to read.

> **Exercise 1.2.** Calculate **(a)** $943 +- 242$, **(b)** $29 \times 31$, **(c)** $\displaystyle \frac{19 + 21}{5 \times 3}$, **(d)**

Hello.
