---
title:  "R Worksheet 1: R Basics (Solutions)"
author: "MATH1710 Probability and Statistcs I"
date:   "University of Leeds, 2021--22"
output: html_document
---

*Here are my solutions to [R Worksheet 1](https://mpaldridge.github.io/math1710/R/R1.html). These aren't necessarily the only ways to solve these problems.*

*Remember that you can get R help at the optional R troubleshooting drop-in sessions in Weeks 2 and 3.*

---

> **Exercise 1.1.** Type `2 + 3` into the console next to the arrow, and then press Enter. What happens?

Let's try!

```{r, eval = FALSE}
2 + 3
```

We see the answer `[1] 5`. First, the `[1]` just tells us that what follows is the first part of the answer -- since our answer here only has one part, we can ignore this. (It is useful when the answer is a very long vector or a table of values.) Then the `5` is the answer we want.

> **Exercise 1.2.** Calculate:\
**(a)** $943 - 242$,\
**(b)** $29 \times 31$,\
**(c)** $2^{8+5}$,\
**(d)** $\displaystyle \frac{19 + 21}{5 \times 3}$.

We calculate these as follows:

```{r}
943 - 242
29 * 31
2^(8 + 5)
(19 + 21) / (5 * 3)
```

> **Exercise 1.3.** Use R to find:\
**(a)** $\frac{1}{7}$ to 4 decimal places;\
**(b)** $\log(10)$ to 3 significant figures;\
**(c)** $\sqrt{712 + 34}$ to the nearest integer.

We can find these as follows:

```{r}
round(1 / 7, digits = 4)
signif(log(10), digits = 3)
round(sqrt(712 + 34))
```

> **Exercise 1.4.** Create an object called `john`, and assign it the value 7. Then create an object called `paul` and assign it the value $12^2$. Then get R to tell you the value of `paul` multiplied by the value of `john`.

We do this as follows:

```{r}
john <- 7
paul <- 12^2
paul * john
```

> **Exercise 1.5**. This exercise continues with the objects assigned in Exercise 1.4.\
**(a)** Assign the value of `paul` mutiplied by `john` to the new value `ringo`.\
**(b)** Check the value of `ringo`.\
**(c)** Double the value of `ringo`, keeping it still stored as `ringo`.\
**(d)** Add 6 to the value of `ringo`.\
**(e)** Check the new value of `ringo`. (It should be 2022.)

We do this as follows:

```{r}
ringo <- paul * john
ringo
ringo <- 2* ringo
ringo <- ringo + 6
ringo
```

The answer is indeed 2022, as it should be.

> **Exercise 1.6.** Write down the commands you used to solve Exercises 1.4 and 1.5 in a new R Script. Save your work with a explanatory filename that will allow you to find it again later.

> **Exercise 1.7.** Continuing with your R Script from Exercise 1.6, add comments to make it clear which commands are doing what, then re-save your R Script.

My R Script `R1-solutions.R` looked like this:

```{r, eval = FALSE}
# MATH1710: R WORKSHEET 1
# MY SOLUTIONS
# Last updated: Friday 30 September 2022

# Exercise 1.1
2 + 3  # Gets output "[1] 5", meaning the answer is 5

# Exercise 1.2
943 - 242
29 * 31
2^(8 + 5)
(19 + 21) / (5 * 3)

# Exercise 1.3
round(1 / 7, digits = 4)
signif(log(10), digits = 3)
round(sqrt(712 + 34))

# Exercise 1.4
john <- 7
paul <- 12^2
paul * john

# Exercise 1.5
ringo <- paul * john
ringo
ringo <- 2* ringo
ringo <- ringo + 6
ringo  # Answer is 2022, as it should be

# Exercise 1.5 created this R Script
# Exercise 1.6 added comments to this R Script
```