---
title:  "R Worksheet 3: Data in R -- Solutions"
author: "MATH1710 Probability and Statistcs I"
date:   "University of Leeds, 2023--24"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

These are my solutions to the exercises on R Worksheet 3. This R code is not the only way to solve these exercises -- although your numerical answers should agree (up to rounding).

> **Exercise 3.1.** Copy the command above to read in the Met Office temperature data to an object called `temperature`. How many rows does the data have?

I started by reading the temperature data into R.

```{r}
temperature <- read.csv("https://mpaldridge.github.io/math1710/data/met-office.csv")
```

I then found the number of rows in the data

```{r}
nrow(temperature)
```


> **Exercise 3.2** We continue with the `temperature` data.\
**(a)** What year does the 40th row correspond to?

```{r}
temperature[40, "year"]
```

> **(b)** What was the temperature in August of that year?

```{r}
temperature[40, "aug"]
```

> **(c)** Output the whole list of January temperatures.

```{r}
temperature$jan
```

> **(d)** Output the December temperature for rows 50 to 60.

```{r}
temperature[50:60, "dec"]
```


> **Exercise 3.3.** We continue with the `temperature` data.\
**(a)** What was the median temperature in September?

```{r}
median(temperature$sep)
```

> **(b)** What is the sample variance of the first fifty years of February data?

```{r}
answer33b <- var(temperature[1:50, "feb"])
round(answer33b, digits = 2)
```

**(c)** What is the lowest December temperature?

```{r}
min(temperature$dec)
```

**(d)** What is the correlation between October and November temperatures, restricted only to the first 100 years of data?

```{r}
answer33d <- cor(temperature[1:100, "oct"], temperature[1:100, "nov"])
signif(answer33d, digits = 2)
```