---
title:  "R Worksheet 3: Data in R"
author: "MATH1710 Probability and Statistcs I"
date:   "University of Leeds, 2022--23"
output: html_document
---

*This R worksheet* does *include assessed questions at the end. Only the question marked "Assessed Question" are assessed, not the other "Exercises". The deadline for submitting your solutions to these questions is **Monday 24 October at 1400**. If you have difficulty with this worksheet, you can get help at your R troubleshooting drop-in in Week 3.*

------------------------------------------------------------------------

## CSV files

If we are dealing with a very small amount of data, then we can simply type it into one or more vectors in R, using the `c()` function. But if we have a large amount of data, we will want to "read" it into R directly from a file -- that is, we will want R to take the data from an external file and turn it into an R object.

When large amounts of data are transferred, the standard format is a **comma-separated variable** or **CSV** file, which typically have the file suffix `.csv`. A comma-separated variable file looks like this:

```
year,jan,feb,mar,apr,may,jun,jul,aug,sep,oct,nov,dec
1884,4.9,3.9,4.9,6.0,9.7,12.6,14.7,15.5,13.0,8.0,4.6,2.9
1885,1.8,4.1,3.2,6.3,7.5,12.4,14.5,12.2,10.7,6.1,4.6,2.7
1886,0.8,0.5,2.6,5.9,8.4,11.6,14.2,14.3,12.0,9.8,5.4,0.9
1887,1.3,3.1,3.0,5.3,8.1,14.1,15.9,13.8,10.4,6.1,3.7,1.8
1888,2.5,1.1,1.7,5.1,9.2,11.3,12.0,12.6,10.8,7.1,6.3,3.9
```

Here, the first line of the file represents a row of "headers", telling us what the rest of the data represents. This is data from the UK Met Office about the average UK temperature in each month of different years. Then each line of the file after the first line corresponds to one data record -- here, one year -- and each piece of data is separated from the next with just a single comma `,` and no spaces. Represented in a more common table view, this data would look like this:

| year | jan | feb | mar | apr | may | jun | jul | aug | sep | oct | nov | dec |
|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|
| 1884 | 4.9 | 3.9 | 4.9 | 6.0 | 9.7 | 12.6 | 14.7 | 15.5 | 13.0 | 8.0 | 4.6 | 2.9 |
| 1885 | 1.8 | 4.1 | 3.2 | 6.3 | 7.5 | 12.4 | 14.5 | 12.2 | 10.7 | 6.1 | 4.6 | 2.7 |
| 1886 | 0.8 | 0.5 | 2.6 | 5.9 | 8.4 | 11.6 | 14.2 | 14.3 | 12.0 | 9.8 | 5.4 | 0.9 |
| 1887 | 1.3 | 3.1 | 3.0 | 5.3 | 8.1 | 14.1 | 15.9 | 13.8 | 10.4 | 6.1 | 3.7 | 1.8 |
| 1888 | 2.5 | 1.1 | 1.7 | 5.1 | 9.2 | 11.3 | 12.0 | 12.6 | 10.8 | 7.1 | 6.3 | 3.9 |

There are many advantages to using CSV files to store and send data:

* They can be read by many different programs: you can open them in a text editor (like Notepad), you can open them in a spreadsheet program (like Excel), or you can read them directly into R.
* Because a CSV file contains just the data and some commas and nothing else, file sizes can be kept small even for very large amounts of data.
* CSV files contain "just the data", and no unnecessary presentational information like fonts, text size, colours, and so on.

We can read data into R from a CSV file saved "locally" -- that is, saved on the hard drive of the computer you are working on now -- or from a CSV file on the internet. In these R worksheets, we will only read in data from the internet: for this module, you never need to save any CSV data files to your computer. The advantages of reading data in from the internet include:

* Your R code will always work on any (internet-connected) computer, for example if you send your work to a colleague.
* If the data owner updates their data on the internet, your R code will work automatically with this new updated data.

A disadvantage is that the data owner might remove their data from the internet.

## Reading CSV files into R

We read data from CSV files into R using the **`read.csv()`** function. To read in more of the UK temperature data we mentioned above from the website <https://mpaldridge.github.io/math1710/data/met-office.csv>, we use the R command

```{r}
temperature <- read.csv("https://mpaldridge.github.io/math1710/data/met-office.csv")
```

By using `temperature <- `, we have read the data into an R object called `temperature`. Nothing will happen until we interact with this R object in some way. (Note the quotation marks `" "` around the web address in the `read.csv()` command.)

*Remember:* to read data from a CSV file you must use the `read.csv()` -- just typing a web address straight into R will not do anything (except return an error).

Once we've read some data into R, we will want to check it has worked, and find out some basic details about the data. There are various functions that will do this:

* **`head()`** shows the first few rows of the data, so you can inspect a manageable amount of the data. (So to show the top of the data read in as `temperature`, we use `head(temperature)`.)
* **`names()`** tells us the names of the columns from the header row of the file.
* **`ncol()`** and **`nrow()`** tell us how many columns and how many rows the data contains.

> **Exercise 3.1.** Copy the command above to read in the Met Office temperature data to an object called `temperature`. How many rows does the data have?

Let us note for later that the names of our `temperature` data are these:

```{r}
names(temperature)
```

## Data frames

When R reads in data from a CSV file, it saves it as a type of object known as a **data frame**. Let us look at the top of `temperature`:

```{r}
head(temperature)
```

A data frame is a two-dimensional table. The columns have names: the names `"year"`, `"jan"`, `"feb"`, etc, that we saw above. The rows are simply numbered.

We can extract a single data point using the square brackets `[ ]`, similar to how we extracted single entries of vectors on Worksheet 2. So, the datapoint in row 21 and column "oct" is

```{r}
temperature[21, "oct"]
```

The notation here in the square brackets is: first the number of the row, then a comma, then the name of the column in quotation marks. (Don't forget the quotation marks!)

You'll also remember from R Worksheet 2 that we can pull out a segment from a column with, for example,

```{r}
temperature[21:26, "oct"]
```

If we want a whole row, we can just omit the column identifier -- but remember that we still need the comma!

```{r}
temperature[21, ]
```

Similarly, if we want a whole column we can miss out the row identifier -- but still keep the comma! -- with

```{r, eval = FALSE}
temperature[, "oct"]
```

(I won't print the whole output here.) However, we want a whole column sufficiently often that there's a shorter and more convenient notation where we just use a **`$`** sign, and don't need to worry about the square brackets, the comma, or the quotation marks.

```{r, eval = FALSE}
temperature$oct
```

> **Exercise 3.2** We continue with the `temperature` data.\
**(a)** What year does the 40th row correspond to?\
**(b)** What was the temperature in August of that year?\
**(c)** Output the whole list of January temperatures.\
**(d)** Output the December temperature for rows 50 to 60.

## Functions with data

Once we extract the rows or columns we need, we can then apply any of the functions we learned about on Worksheet 2, like `mean()`, `median()`, `var()`, `IQR()`, `cor()`, `min()`, `max()` and so on.

So, for example, the mean February temperature is given by

```{r}
mean(temperature$feb)
```

the interquartile range of April temperatures in the first 100 rows is given by

```{r}
IQR(temperature[1:100, "apr"])
```

the correlation between June and July temperatures is

```{r}
cor(temperature$jun, temperature$jul)
```

and so on. Remember we can use `round()` and `signif()` with these to round the answers.

> **Exercise 3.3.** We continue with the `temperature` data.\
**(a)** What was the median temperature in September?\
**(b)** What is the sample variance of the first fifty years of February data?\
**(c)** What is the lowest December temperature?\
**(d)** What is the correlation between October and November temperatures, restricted only to the first 100 years of data?

---

## Assessed questions

*The following five assessed questions should be submitted to the Microsoft Form that is linked to below. I recommend you do this in Week 3, but the official deadline is **Monday 24 October at 1400**.*

*You should submit your work by typing your answers into [this Microsoft Form](https://forms.office.com/Pages/ResponsePage.aspx?id=qO3qvR3IzkWGPlIypTW3ywARQdZlKXRHsLcXi_ngX8NURUJBT080Mks5NkVNRzhMQldDMFBLNUkxVy4u). You may need to be logged into your University account.*

*This worked will be marked automatically by computer, so make sure you answers are accurate -- the computer does not "know what you meant"; only what you actually enter into the form.*

*So that (most) students get different data to work with, you will work from a data file whose name is based on your student ID number. You student ID number is (usually) a 9-digit number starting `201`. Your data set is the CSV file at*

```
https://mpaldridge.github.io/math1710/data/R3-xy.csv
```

*where `xy` is replaced by the last two digits of your student ID number. So if your student ID is `201623429`, then you should use the file at `https://mpaldridge.github.io/math1710/data/R3-29.csv`; if your student ID number is `201491200` then you should use the file `https://mpaldridge.github.io/math1710/data/R3-00.csv` and so on. Take care to check you get this correct: if you get your student ID number wrong and/or use the wrong data file, the computer is likely to award you 0 total marks.*

*If asked to round your answer, you should give your answer as produced by R's `round()` or `signif()` functions.*

---

**First**, read the data set at `https://mpaldridge.github.io/math1710/data/R3-xy.csv` into R. Remember that `xy` should be replaced by the last 2 digits of your student ID number. This data set gives the stock prices of two companies, "Amaflix" and "Netazon" over a number of days.

> **Assessed Question 1.** How many days of data are in the data set?

> **Assessed Question 2.** What was the mean stock price of Amaflix? Give your answer to two decimal places.

> **Assessed Question 3.** What was the variance of stock price of Netazon over the first 25 days of the dataset? Round your answer to three significant figures.

> **Assessed Question 4.** Which company's data had the greater inter-quartile range?

> **Assessed Question 5.** Would you say that the relationship between the companies' stock prices is a strong positive relationship, a strong negative relationship, or a weak relationship?

---

*Remember to submit your answers into  [this Microsoft Form](https://forms.office.com/Pages/ResponsePage.aspx?id=qO3qvR3IzkWGPlIypTW3ywARQdZlKXRHsLcXi_ngX8NURUJBT080Mks5NkVNRzhMQldDMFBLNUkxVy4u) before 1400 on Monday 24 October.*