---
title:  "R Worksheet 4: Plots I: Making plots"
author: "MATH1710 Probability and Statistcs I"
date:   "University of Leeds, 2023--24"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = NA)
```


> **Exercise 4.1.** Read the temperature data into R by copying the command above. Just to check you remember from R Worksheet 3: What was the median August temperature?

We start by reading in the Met Office data using the `read.csv()` function.

```{r}
temperature <- read.csv("https://mpaldridge.github.io/math1710/data/met-office.csv")
```

The median August temperature is

```{r}
median(temperature$aug)
```

> **Exercise 4.2.** Draw a boxplot of the August temperature. Save your graph as an image.

```{r}
boxplot(temperature$aug)
```

I saved the graph as an image by clicking on "Export" in the "Plots" window in RStudio, then choosing "Save as Image...".

> **Exercise 4.2.** Draw a boxplot comparing the April temperature for the first 68 years in the data set (rows 1 to 68) and the second 68 years in the data (rows 69 to 136). (You may need to look back at R Worksheet 3 to remind yourself how to extract limited rows from a column of a data frame.) Comment briefly on what you learn from the comparison.

```{r}
boxplot(temperature[1:68, "apr"], temperature[69:136, "apr"])
```

I found that XXXXXXXX YYYYYYY ZZZZZZZZZ WWWWWWWW.

```{r}
hist(temperature$dec, freq = FALSE)
```

> **Exercise 4.4.** Plot four histograms of the May temperature with density on the y-axis using (approximately) 3, 6, 10, and 20 bins. Which number of bins do you think is the most appropriate, and why?



> **Exercise 4.5.** Plot a scatterplot for January and August temperature. Do you see a relationship here? Copy your scatterplot to the clipboard and paste it into some other program.



> **Exercise 4.6.** Draw a scatterplot for the first 30 years of data, with year as the x-axis and February temperature as the y-axis, with points plotted that are also joined up with lines.