---
title:  "R Worksheet 8: Discrete random variables"
author: "MATH1710 Probability and Statistcs I"
date:   "University of Leeds, 2021--22"
output: html_document
---

*This R worksheet does not include assessed questions.You should complete this worksheet in Week 8. If you have difficulty with this worksheet, you can get help at the office hours drop-in sessions.*

------------------------------------------------------------------------

On the last worksheet, we looked at dealing with famous discrete random variables: the binomial, geometric, and Poisson distributions. In this worksheet, we will look at dealing with arbitrary discrete random variables, as defined by a probability mass function (PMF)

## Setting up a PMF

The key to dealing with a discrete random variable is to set up two things: the range of values the random variable can take, and the PMF itself.

Consider, for example, the following PMF for a random variable $X$:

|   $x$    |  $0$  | $0.5$ |  $1$  |  $2$  | $2.5$ |  $4$  |
|:--------:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| $p_X(x)$ | $0.1$ | $0.2$ | $0.2$ | $0.3$ | $0.1$ | $0.1$ |

We want to store the $x$ values in one vector, and the $p_X(x)$ values in another vector. We can do that like this:

```{r}
x <- c(0, 0.5, 1, 2, 2.5, 4)
pmf_x <- x(0.1, 0.2, 0.2, 0.3, 0.1, 0.1)
```

It's good practice to check that the PMF does indeed sum to 1:

```{r}
sum(pmf_x)
```