---
title:  "R Worksheet 8: Discrete random variables"
author: "MATH1710 Probability and Statistcs I"
date:   "University of Leeds, 2021--22"
output: html_document
---

*This R worksheet does not include assessed questions.You should complete this worksheet in Week 8. If you have difficulty with this worksheet, you can get help at the office hours drop-in sessions.*

------------------------------------------------------------------------

On the last worksheet, we looked at dealing with famous discrete random variables: the binomial, geometric, and Poisson distributions. In this worksheet, we will look at dealing with arbitrary discrete random variables, as defined by a probability mass function (PMF)

## Probability mass function

The key to dealing with a discrete random variable is to set up two things: the range of values the random variable can take, and the PMF itself.

Consider, for example, the following PMF for a random variable $X$:

|   $x$    |  $0$  | $0.5$ |  $1$  |  $2$  | $2.5$ |  $4$  |
|:--------:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| $p_X(x)$ | $0.1$ | $0.2$ | $0.2$ | $0.3$ | $0.1$ | $0.1$ |

We want to store the $x$ values in one vector, and the $p_X(x)$ values in another vector. We can do that like this:

```{r}
x <- c(0, 0.5, 1, 2, 2.5, 4)
pmf_x <- c(0.1, 0.2, 0.2, 0.3, 0.1, 0.1)
```

It's good practice to check that the PMF does indeed sum to 1:

```{r}
sum(pmf_x)
```

We can find the probability of a given value of $x$ by pulling out the relevant value of the PMF. We are used to using square brackets to pull out the elements of a vector. But more complicate expressions can be put inside the brackets, like this:

* `pmf_x[x == 0.5]` pulls out the value of the PMF that corresponds to $x = 0.5$; that is, of $p_X(0.5)$. (Note the double-equals `==`. Roughly speaking, in R, a single-equals `=` means "is set to be equal to", while a double-equals `==` means "*is* equal to".)
* `pmf_x[x > 1.5]` pulls out all the values of the PMF that corresponding to $x > 1.5$. It's often helpful to use this together with the `sum()` function, because `sum(pmf_x[x > 1.5])` will give $\mathbb P(X > 1.5)$. (Do you see why?)
* Relations that can be used this way are
   * `==`: equals $=$
   * `>`: greater than $>$
   * `<`: less than $>$
   * `>=`: greater than or equal to $\geq$
   * `<=`: less than or equal to $\leq$
   * Keen R users may wish to research use of `!=`, `in`, `&`, `|`.
   
> **Exercise 8.1.** Let $Y$ be the sum of fair dice rolls. So $Y$ takes the values $(2, 3, \dots, 12)$ with probabilities $\frac{1}{36}(1,2,3,4,5,6,5,4,3,2,1)$.\
**(a)** Enter this random variable into `y` and `pmf_y`. Do so as efficiently as possible, by taking advantage of the colon notation.\
**(b)** What is $\mathbb P(Y = 9)$?\
**(c)** What is $\mathbb P(Y \leq 6)$?

## Plotting a PMF

## Cumulative distribution function

## Expectation

## Sampling from a distribution
